---
title: "Ensemble TADPOLE leaderboard forecasts for accuracy improvement"
author: "Jonas Richiardi"
date: "3/17/2019"
output: html_document
---

# INTRO

This is an baseline for combining several TADPOLE Leaderboard forecasts, initially used during the BrainHack Open Geneva <http://brainhack.ch> for the Dementia forecasting machine learning project.

For information on TADPOLE please look at <https://tadpole.grand-challenge.org>

Usage of the ADNI data present in the TADPOLE challenge,is governed by the ADNI data usage and citation policy.

# PACKAGE LOADING AND DATA PREP

```{r setup_datprep, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# General programming constructs
library(plyr)
library(tidyverse)
library(glue)

# General computation
library(doParallel)

# Viz
library(visdat)
library(directlabels)

# where the TADPOLE challenge forecasts csv files are
results_path<-'../../results'

all_forecast_files<-dir(results_path, pattern="*.csv")

df_fc<-plyr::ldply(all_forecast_files, .fun=function(this_file){
  df_fc_tmp<-readr::read_csv(file.path(results_path,this_file))
  return(cbind(data.frame(ForecastFile=this_file), df_fc_tmp))
})

# quick look
ggplot(df_fc, aes(`CN relative probability`))+
  geom_density(aes(colour=ForecastFile))+
  guides(colour=FALSE)+
  ggitle('Distribution per forecast file')

# check out one subject
filter(df_fc, RID==8, `Forecast Date`=="2012-01")



```


# ENSEMBLE FORECASTS

## Averaging

```{r ensemble_forecasts_avg}

df_ens<-dplyr::select(df_fc, -ForecastFile) %>% dplyr::group_by(RID,`Forecast Month`,`Forecast Date`) %>% dplyr::summarise_if(is.numeric, mean, na.rm = TRUE) %>% ungroup()

# check out a subject post-ensembling
filter(df_ens, RID==8, `Forecast Date`=="2012-01")


```

## Median 

```{r ensemble_forecasts_median}

df_ens<-dplyr::select(df_fc, -ForecastFile) %>% dplyr::group_by(RID,`Forecast Month`,`Forecast Date`) %>% dplyr::summarise_if(is.numeric, median, na.rm = TRUE) %>% ungroup()

# check out a subject post-ensembling
filter(df_ens, RID==8, `Forecast Date`=="2012-01")


```


# PREPARE SUBMISSION FILE 

```{r prepare_sub}

# submission name - no spaces, no special chars, no funny business
name_submission<-'Ensemble_median'
output_file = file.path(results_path, glue::glue('TADPOLE_Submission_Leaderboard_{name_submission}.csv'))

readr::write_csv(df_ens, output_file)


```

